---
title: R color scheme
author: Ming
date: '2022-09-01'
slug: []
categories:
  - R
tags:
  - data visualization
  - ggplot
subtitle: ''
lastmod: '2023-05-26T18:31:23-07:00'
authorLink: ''
description: ''
license: ''
images: []
featuredImage: ''
featuredImagePreview: ''
hiddenFromHomePage: no
hiddenFromSearch: no
twemoji: no
lightgallery: yes
ruby: yes
fraction: yes
fontawesome: yes
linkToMarkdown: yes
rssFullText: no
toc:
  enable: yes
  auto: yes
code:
  copy: yes
  maxShownLines: 50
math:
  enable: no
mapbox: ~
share:
  enable: yes
comment:
  enable: yes
library:
  css: ~
  js: ~
seo:
  images: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=F,warning = F,cache=F)
library(ggplot2)
```

# Discrete colors

## 3-color palette

```{r fig.height=3}
# Box plot
bp<-ggplot(ToothGrowth, aes(x=factor(dose), y=len, fill=factor(dose))) +
  geom_boxplot()+theme_bw()
bp + scale_fill_manual(values=c("#999999", "#E69F00", "#56B4E9"))

# Scatter plot
sp<-ggplot(mtcars, aes(x=factor(wt), y=mpg, color=factor(cyl))) + geom_point(size=4)+theme_bw()
sp + scale_color_manual(values=c("#999999", "#E69F00", "#56B4E9"))
```

## another 3-color palette

```{r fig.height=3}
# Box plot
bp<-ggplot(ToothGrowth, aes(x=factor(dose), y=len, fill=factor(dose))) +
  geom_boxplot()+theme_bw()
bp + scale_fill_manual(values = c("#00AFBB", "#E7B800", "#FC4E07"))

# Scatter plot
sp<-ggplot(mtcars, aes(x=factor(wt), y=mpg, color=factor(cyl))) + geom_point(size=4)+theme_bw()
sp + scale_color_manual(values = c("#00AFBB", "#E7B800", "#FC4E07"))

```

## colorblind-friendly palette

```{r fig.height=3}
custom.col <- c("#FFDB6D", "#C4961A", "#F4EDCA", 
                "#D16103", "#C3D7A4", "#52854C", "#4E84C4", "#293352")
barplot(rep(1,length(custom.col)),col=custom.col)

# The palette with grey:
cbp1 <- c("#999999", "#E69F00", "#56B4E9", "#009E73",
          "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
barplot(rep(1,length(cbp1)),col=cbp1)
# The palette with black:
cbp2 <- c("#000000", "#E69F00", "#56B4E9", "#009E73",
          "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
barplot(rep(1,length(cbp2)),col=cbp2)
```

## ggplot2 default color palette
refer to: https://stackoverflow.com/questions/8197559/emulate-ggplot2-default-color-palette
```{r fig.height=3}
library(scales)
hue_pal()(4)
#[1] "#F8766D" "#7CAE00" "#00BFC4" "#C77CFF"
show_col(hue_pal()(4))
show_col(hue_pal()(3))
```

## RColorBrewer package
refer to: https://rdrr.io/cran/RColorBrewer/man/ColorBrewer.html

```{r fig.height=3}
library(RColorBrewer)
brewer.pal.info
#par(mar=c(3,4,2,2))
#display.brewer.all()
(mycol=brewer.pal(10,"Paired"))
barplot(rep(1,10),col=mycol)
(mycol=brewer.pal(8,"Dark2"))
barplot(rep(1,8),col=mycol)
```

# Continuous colors

## scale_xxx_gradient()
frequently used in heatmaps
refer to: http://www.sthda.com/english/wiki/ggplot2-colors-how-to-change-colors-automatically-and-manually#continuous-colors

The graph can be colored according to the values of a continuous variable using the functions :

- scale_color_gradient(), scale_fill_gradient() for sequential gradients between two colors
- scale_color_gradient2(), scale_fill_gradient2() for diverging gradients
- scale_color_gradientn(), scale_fill_gradientn() for gradient between n colors


```{r fig.height=3}
# Color by qsec values
sp<-ggplot(mtcars, aes(x=wt, y=mpg, color=qsec)) + geom_point() +theme_bw()
#sp
summary(mtcars$qsec)
sp+scale_color_gradient2(low="blue", mid = "white",high="red",
        midpoint = quantile(mtcars$qsec,0.25))
```

## viridis package
useful refs: 

- https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html
- https://www.thinkingondata.com/something-about-viridis-library/
- https://sjmgarnier.github.io/viridisLite/reference/viridis.html
- https://bookdown.org/hneth/ds4psy/D-4-apx-colors-pkgs.html

```{r fig.height=3}
library(viridis)
ggplot(data.frame(x = rnorm(10000), y = rnorm(10000)), aes(x = x, y = y))+
  geom_hex() + coord_fixed() +
  scale_fill_viridis() + theme_bw()
```

It's also easier to adjust continuous colors into discrete ones with the `viridis` package.

```{r fig.height=3}
p <- ggplot(mtcars, aes(wt, mpg))+
    geom_point(size=4, aes(colour = factor(cyl))) +
    theme_bw()
p+scale_color_viridis(discrete=TRUE,option = 'viridis') 
p+scale_color_viridis(discrete=TRUE,option = 'magma') 
p+scale_color_viridis(discrete=TRUE,option = 'turbo') 
```

```{r}
library(viridisLite)  # load package
library(unikn) #seecol function
vir_10 <- viridis(n = 10)
seecol(vir_10, col_brd = "white", lwd_brd = 4, 
       title = "Example of a viridis color palette (n = 10)",
       pal_names = paste0("viridis_", 1:10))
```

# Other R color combos that are colorblind-friendly

## 8-color palette
```{r}
colorBlindBlack8  <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
                       "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
pie(rep(1, 8), col = colorBlindBlack8)

colorBlindGrey8   <- c("#999999", "#E69F00", "#56B4E9", "#009E73", 
                       "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
scales::show_col(colorBlindGrey8)
```

## 12-color palette
```{r}
safe_colorblind_palette <- c("#88CCEE", "#CC6677", "#DDCC77", "#117733", "#332288", "#AA4499", 
                             "#44AA99", "#999933", "#882255", "#661100", "#6699CC", "#888888")
scales::show_col(safe_colorblind_palette)
```

## RColorBrewer: display.brewer.all(colorblindFriendly = T)
refer to: [colorblind-friendly-palettes](https://riptutorial.com/r/example/28354/colorblind-friendly-palettes)
```{r fig.height=8}
library(RColorBrewer)
display.brewer.all(colorblindFriendly = T)
#palette using grey
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
scales::show_col(cbPalette)
#palette using black
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
scales::show_col(cbPalette)
```

# Miscellaneous
## make transparent colors in R
refer to: [Make transparent colors in R](https://www.dataanalytics.org.uk/make-transparent-colors-in-r/#:~:text=Make%20transparent%20colors%20in%20R&text=The%20rgb()%20command%20is,255%20being%20%E2%80%9Csolid%E2%80%9D)

```{r}
## Transparent colors
## Mark Gardener 2015
## www.dataanalytics.org.uk

t_col <- function(color, percent = 50, name = NULL) {
  #      color = color name
  #      percent = % transparency
  #      name = an optional name for the color
  
  ## Get RGB values for named color
  #rgb.val <- col2rgb(color)
  
  rgb.val=col2rgb(color)
  
  ## Make new color using input color as base and alpha set by transparency
  t.col <- rgb(rgb.val[1], rgb.val[2], rgb.val[3],
               max = 255,
               alpha = (100 - percent) * 255 / 100,
               names = name)
  
  ## Save the color
  #invisible(t.col)
  t.col
}
my.cols=c('black','orange','sky blue','yellow','blue');
my.cols
my.cols2=sapply(my.cols,function(i){t_col(i,percent=80)})
my.cols2
par(mfrow=c(2,1))
barplot(1:length(my.cols),col=my.cols)
barplot(1:length(my.cols),col=my.cols2)
```

# Useful resources

1. [sthda: how to change colors automatically and manually](http://www.sthda.com/english/wiki/ggplot2-colors-how-to-change-colors-automatically-and-manually)
2. [GGPlot Colors Best Tricks You Will Love by Alboukadel](https://www.datanovia.com/en/blog/ggplot-colors-best-tricks-you-will-love/)
3. [stackoverflow: r-plot-color-combinations-that-are-colorblind-accessible](https://stackoverflow.com/questions/57153428/r-plot-color-combinations-that-are-colorblind-accessible)
4. [color-safe-palett](https://mikemol.github.io/technique/colorblind/2018/02/11/color-safe-palette.html)
5. [tokyo.ac.jp](http://jfly.iam.u-tokyo.ac.jp/color/#pallet)

```{r}
#installed.packages()[names(sessionInfo()$otherPkgs), "Version"]
sessionInfo()
```


