---
title: Linear Regression - log-log plot and fit a line
author: Ming
date: '2022-04-06'
slug: []
categories:
  - statistics
tags:
  - linear regression
  - statistics
  - data visualization
subtitle: ''
lastmod: '2023-05-26T15:05:15-07:00'
authorLink: ''
description: ''
license: ''
images: []
featuredImage: ''
featuredImagePreview: ''
hiddenFromHomePage: no
hiddenFromSearch: no
twemoji: no
lightgallery: yes
ruby: yes
fraction: yes
fontawesome: yes
linkToMarkdown: yes
rssFullText: no
toc:
  enable: yes
  auto: yes
code:
  copy: yes
  maxShownLines: 50
math:
  enable: no
mapbox: ~
share:
  enable: yes
comment:
  enable: yes
library:
  css: ~
  js: ~
seo:
  images: []
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T,message = F,warning = F,
        fig.width=8,fig.height=4,fig.align='center')
library(knitr)
library(tidyverse)
```

## log-log plot is "just" a way to "squeeze" the plot

In a log-log plot, 
we are seeing the relationship between `log(x)` and `log(y)`

```{r}
# polynomial, generate some data
x<-1:10; y=x*x
# Simple graph
plot(x, y)

# Log on y-scale alone
par(mfrow=c(1,2),mar=c(5,4,2,1))
plot(x, y, log="y")
plot(x,log(y))
# distribution shapes are the same, just y-axis labeled differently.

# Log both axises
plot(x,y,log="xy")
plot(log(x),log(y))
```

Why it is a straight line?

Remember we are looking at `log(x)` and `log(y)` on the log-log plot.

Let's rephrase $y = x^{2}$ by taking logarithm on both sides:

$log(y) = 2 * log(x)$

So, exactly, we expect to see a straight line with intercept 0 and slope 2.

Let's try fit a regression line and check the slope.


## but you could fit simple linear regression lines now!
```{r}
new.y=log(y)
new.x=log(x)
out<-lm(new.y~new.x)
out
plot(log(x),log(y))
abline(out)
```

## have fun with exponential distribution
$f(x) = a*e^{-bx}$

$log(f(x)) = log(a) - bx$

Let's make thelog-log plot and see if there is a linear relationship.

```{r}
# exponential, generate some data
a=3;
b=2;
x=1:10
y=a*exp(-b*x)
par(mfrow=c(1,3))
plot(x,y)
plot(x,y,log='xy')
plot(log(x),log(y))

(out=lm(log(y)~log(x)))
```


Wait.... The middle and rightmost plots do not look like a straight line.

What happened?

Remember, on a log-lot plot we are seeing the relationship between `log(y)` and `log(x)`, **NOT** `log(y)` and `x`.


So what's the relationship between `log(y)` and `log(x)` after all.

Let's "play magic" on the equation.

While there is no explict `log(x)` in the above equation,

we can make it happen by the fact: $x = exp^{log(x)}$,

then we have:

$log(f(x)) = log(a) - b*exp^{log(x)}$

So, it's pretty clear we don't have a linear relationship between `log(x)` and `log(y)`.

Sad... but below is something fun!

## Power Function and Power Law distribution

I won't talk too much about the Power Law distribution as it's too cheap and easy and cliche... It's not cool!!!

Just a snippet.
$y = Cx^{-a}$

$log(y) = log(C) - a*log(x)$

```{r}
C=3
a=2
x=1:10
y=C*x^(-a)
par(mfrow=c(1,3))
plot(x,y)
plot(x,y,log='xy')
plot(log(x),log(y))

(out<-lm(log(y)~log(x)))
log(C)
-a
plot(log(x),log(y))
abline(out)
```

