---
title: 'Linear Regression: Visulaze & Interpret the Interaction'
author: Ming
date: '2022-04-05'
slug: []
categories:
  - statistics
tags:
  - statistics
  - linear regression
subtitle: ''
lastmod: '2023-05-26T14:07:59-07:00'
authorLink: ''
description: ''
license: ''
images: []
featuredImage: ''
featuredImagePreview: ''
hiddenFromHomePage: no
hiddenFromSearch: no
twemoji: no
lightgallery: yes
ruby: yes
fraction: yes
fontawesome: yes
linkToMarkdown: yes
rssFullText: no
toc:
  enable: yes
  auto: yes
code:
  copy: yes
  maxShownLines: 50
math:
  enable: no
mapbox: ~
share:
  enable: yes
comment:
  enable: yes
library:
  css: ~
  js: ~
seo:
  images: []
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T,message = F,warning = F,
        fig.width=5,fig.height=3,fig.align='center')
library(knitr)
```

## basic model
$y = b0 + b1*x1 + b2*x2 + b3*(x1+x2)$

## Scenario 1: one numeric + one factor

### load the data
```{r}
library(tidyverse)
data(mtcars)
glimpse(mtcars)
# to start with, visualize each predictor effect separately
par(mfrow=c(1,2),mar=c(3,3,1,1))
plot(mtcars$disp,mtcars$mpg)
boxplot(mpg~vs,data=mtcars)
# try to have a holistic view
ggplot(data=mtcars,aes(x=disp,y=mpg))+geom_point()+facet_wrap(.~vs)+theme_bw(base_size = 15)
# make it even simple and clear
ggplot(data=mtcars,aes(x=disp,y=mpg,col=factor(vs)))+geom_point()+theme_bw(base_size = 15)
```

Asking 'is there interaction between `disp` and `vs`', is like asking: if I fit lines for green dots and red dots separately, 

- would the two lines be the same? 
- have different intercept? 
- have different slopes?

### build two linear models
```{r}
m0<-lm(mpg~disp+vs,data=mtcars)
m1<-lm(mpg~disp*vs,data=mtcars)
anova(m0,m1)
```
The `anova` shows linear model with the interaction term, `m1` in the above code, fits the data slightly better with a pvalue 0.01335.

Then how to interpret the `m0` and `m1` models?    

As the basic linear model is: $y = b0 + b1*x1 + b2*x2 + b3*(x1+x2)$    
in our case, x1=disp, x2=vs


### interprete the model without interaction
for model `m0`: $y = b0 + b1*disp + b2*vs$    

- **vs=0**, $y = b0 + b1*disp$
- **vs=1**, $y = b0 + b1*disp + b2$
<br>   or, $y = b0 +b2 + b1*disp$

Thus, without the interaction term, the effect of `vs` only show up for the intercept, which means there are two parallele lines for blue and red dots.

```{r}
(x=coef(m0))
ggplot(data=mtcars,aes(x=disp,y=mpg,col=factor(vs)))+geom_point()+theme_bw(base_size = 15)+
  geom_abline(intercept = x[1], slope = x[2], color="#F8766D", 
                 linetype="dashed", size=1)+ 
  geom_abline(intercept = x[1]+x[3], slope = x[2], color="#00BFC4", 
                 linetype="dashed", size=1)
```  

To interpret `m0`,     

- holding `vs` constant, if increase `disp` by 100, the `mpg` would increase b1\*100 = `r x[2]*100`.
- holding `disp` constant, cars with `vs=1` is b2 = `r x[3]` larger than `vs=0` ones.

### interprete the model with interaction

for model `m1`: $y = b0 + b1*disp + b2*vs + b3*disp*vs$   

- **vs=0**, $y = b0 + b1*disp$
- **vs=1**, $y = b0 + b1*disp + b2 + b3*disp$
 <br>   or, $y = b0 +b2 + (b1+b3)*disp$

Thus, with the interaction term, the effect of `vs` show up both for the intercept and the slope, which means the lines would intercross.

```{r}
(x=coef(m1)) #b0,b1,b2,b3
ggplot(data=mtcars,aes(x=disp,y=mpg,col=factor(vs)))+geom_point()+theme_bw(base_size = 15)+
  geom_abline(intercept = x[1], slope = x[2], color="#F8766D", 
                 linetype="dashed", size=1)+ 
  geom_abline(intercept = x[1]+x[3], slope = x[2]+x[4], color="#00BFC4", 
                 linetype="dashed", size=1)
```  

To interpret `m1`,     

- holding `vs` constant, if increase `disp` by 100, 
  + if `vs=0`, the `mpg` would increase b1\*100 = `r x[2]*100`.
  + if `vs=1`, the `mpg` would increase (b1+b3)\*100 = `r (x[2]+x[4])*100`.

- holding `disp` constant, cars with `vs=1` is b2+b3\*disp = `r x[3]` + `r x[4]`\*disp larger than `vs=0` ones.

## Scenario 2: one factor + one factor

Basic linear model: $y = b0 + b1*x1 + b2*x2 + b3*(x1+x2)$    
in this case, x1=am, x2=vs

```{r}
# to start with, visualize each predictor effect separately
par(mfrow=c(1,2),mar=c(3,3,1,1))
boxplot(mpg~am,data=mtcars)
boxplot(mpg~vs,data=mtcars)
# try to have a holistic view
ggplot(data=mtcars,aes(x=factor(am),y=mpg))+geom_boxplot()+facet_wrap(.~vs)+theme_bw(base_size = 15)
# make it even simple and clear
ggplot(data=mtcars,aes(x=factor(am),y=mpg,col=factor(vs)))+geom_boxplot()+theme_bw(base_size = 15)
ggplot(data=mtcars,aes(x=factor(am),y=mpg,col=factor(vs)))+
  geom_line(aes(group=factor(vs)),position = position_dodge(width = 0.5))+geom_point(position = position_dodge(width = 0.5))+theme_bw(base_size = 15)

m0=lm(mpg~vs+am,data=mtcars)
m1=lm(mpg~vs*am,data=mtcars)
anova(m0,m1)
```

**without interaction** $y = b0 + b1*am + b2*vs$    

- **vs=0**, $y = b0 + b1*am$
- **vs=1**, $y = b0 + b2 + b1*am$

Similarly, 

- **am=0**, $y = b0 + b2*vs$
- **am=1**, $y = b0 + b1 + b2*vs$

The effect of `am` or `vs` does not depend on the other term.

- Holding `vs` constant, increase of 100 `am` lead to b1*100 increase in `mpg`.
- Holding `am` constant, increase of 100 `vs` lead to b2*100 increase in `mpg`.



**with interaction** $y = b0 + b1*am + b2*vs + b3*am*vs$    

- **vs=0**, $y = b0 + b1*am$
- **vs=1**, $y = b0 + b2 + (b1+b3)*am$

Similarly, 

- **am=0**, $y = b0 + b2*vs$
- **am=1**, $y = b0 + b1 + (b2+b3)*vs$

Factors of `am` or `vs` affect each other.

- Holding `vs` constant, increase of 100 `am` lead to (b1+b3)*100 increase in `mpg`.
- Holding `am` constant, increase of 100 `vs` lead to (b2+b3)*100 increase in `mpg`.

## Scenario 3: one numeric + one numeric
```{r}
# to start with, visualize each predictor effect separately
par(mfrow=c(1,2),mar=c(3,3,1,1))
plot(mpg~disp,data=mtcars)
plot(mpg~drat,data=mtcars)
# try to have a holistic view, as it's two numeric predictor,
# there is no easy to visualize them
ggplot(data=mtcars,aes(x=disp,y=mpg,col=drat))+geom_point()+theme_bw(base_size = 15)

m0=lm(mpg~disp+drat,data=mtcars)
m1=lm(mpg~disp*drat,data=mtcars)
anova(m0,m1)
```

### interprete the model without interaction

**without interaction** $y = b0 + b1*disp + b2*drat$    

The effect of `disp` or `drat` does not depend on the other term.

- Holding `drat` constant, increase of 100 `disp` lead to b1*100 increase in `mpg`.
- Holding `disp` constant, increase of 100 `drat` lead to b2*100 increase in `mpg`.

```{r}
(x=coef(m0))
g1<-ggplot(data=mtcars,aes(x=disp,y=mpg))+geom_point()+theme_bw(base_size = 15)+xlim(c(0,500))+
  geom_abline(intercept = x[1], slope = x[2], color="#F8766D", 
                 linetype="dashed", size=1)
g2<-ggplot(data=mtcars,aes(x=drat,y=mpg))+geom_point()+theme_bw(base_size = 15)+xlim(c(0,5))+
  geom_abline(intercept = x[1], slope = x[3], color="#00BFC4", 
                 linetype="dashed", size=1)
gridExtra::grid.arrange(g1,g2,ncol=2)
```  


### interprete the model with interaction

**without interaction** $y = b0 + b1*disp + b2*drat + b3*disp*drat$    

The effect of `disp` or `drat` depends on the other term.

- Holding `drat` constant, increase of 100 `disp` lead to (b1+b3\*drat)*100 increase in `mpg`.
- Holding `disp` constant, increase of 100 `drat` lead to (b2+b3\*disp)*100 increase in `mpg`.

<br>
**Some useful ref**
https://ademos.people.uic.edu/Chapter13.html#2_continuous_x_continuous_regression



